---
title: "Kruger_Rodents"
format: html
---

```{r}
library(foreach)
library(doParallel)

mpath <- ''
source('clarkFunctions2025.r')
source('captureRecaptureFunctions.R')
```

```{r}
load( 'rodentDataAll.rdata', verbose = T )

# aggregate some related species with low numbers individually
y$commonName[ grep( 'rock rat', y$commonName ) ] <- 'rock rat'
y$commonName[ grep( 'multimammate mouse', y$commonName ) ] <- 'multimammate mouse'

specTab   <- table( y$plot, y$commonName)
specSum   <- colSums( specTab )
specNames <- names( specSum )[ specSum > 4 ]
nspec   <- length( specNames )

specTab
```

## **Model elements**

There will be an estimate of population size for each species 洧맙 at each plot 洧녲k in each of two years 洧노t. Subscripts are needed to keep this straight. The estimate of population size is a **posterior distribution** for each species/plot/year, 洧녜洧,洧녲,洧노Ns,k,t. There is uncertainty in this estimate that can be described by a **standard error (SE)** and **credible interval (CI)**. The SE is the standard deviation of the posterior distribution. The CI is the interior 95% of this distribution with 2.5% in the lower and upper tails.

Again, the potential population size is selected to be larger than the number of individuals observed but still reasonable. If 洧녵洧,洧녲,洧노ns,k,t individuals are observed, then I choose 洧洧,洧녲,洧노=8칑洧녵洧,洧녲,洧노Ms,k,t=8칑ns,k,t. The scaling factor is `n2M` = 8 in this example.

The code used for the analysis uses Markov chain Monte Carlo (MCMC) to simulate a posterior distribution of estimates. Here are the objects it will generate:

|                |                                        |
|----------------|----------------------------------------|
|                | Objects in this analysis               |
| objects        | description                            |
| N              | number of individuals                  |
| beta_0, beta_1 | movement kernel parameters             |
| p              | detection probability                  |
| M              | potential pop size                     |
| psi            | membership probability (fraction of M) |
| movement scale | meters                                 |

Each estimate will come in the form of MCMC chains, which can be combined to approximate marginal distributions for each parameter.

## **Model fitting**

As mentioned above, model fitting is by MCMC, which draws samples that are held as chains of values. It may be easiest to see this with an example. In the block below I set up some objects needed for the analysis:

```{r}
n2M    <- 8                                      # scaling factor
n.mcmc <- 200                                  # no. MCMC iterations
n.sim  <- 1000                                   # no. simulation locations for movement
n.p    <- 100                                    # length of distance vector
d.vec  <- seq(0, 40, length = n.p)               # distance vector for movement

mu.beta   <- c(-2,-.002)                         # prior mean
beta.tune <- c(.3,.0005)                         # SD for beta proposals (MCMC stuff)
s2.beta   <- c( .5, .4 )                         # prior variance

fh       <- c( 'estimate', 'SE', 'CI_025', 'CI_975' )                      # includes 95% CI
bnames   <- c(paste( 'B1', fh, sep = '.' ), paste( 'B2', fh, sep = '.' ) ) # beta estimates
nnames   <- paste( 'N', fh, sep = '.' )                                    # population size
pnames   <- paste( 'P', fh, sep = '.' )                                    # detection Pr
qnames   <- paste( 'Psi', fh, sep = '.' )                                  # membership Pr
dnames   <- paste( 'L', fh, sep = '.' )                                    # length scale
capTimes <- paste( 'c', c(1:4), sep = '')                                  # times recaptured
```

Ordinarily I would set `n.mcmc` much larger than in this example, which executes quickly for illustration.

## **Model for one species**

Here I fit the model for one species/plot/year. I need to extract from `data.frame y` a 洧칑洧녵L칑n matrix 洧녧Y for each species-plot-year. I set up model elements and then generate maps for the grid showing capture locations for each individual.

```{r}
plot <- "LTGran-13R"
yj <- y[ y$commonName == "multimammate mouse" & y$plot == plot & y$year == 2024, ] # subset
yt <- table( yj$trap, yj$id )                                                      # captures 
id <- colnames( yt )                                                               # individuals
n  <- length( id )
L  <- nrow( X )
Y  <- matrix( 0, L, n, dimnames = list( rownames(X), id ) )          # where each ID caught
Y[ rownames( yt ), colnames( yt ) ] <- yt                            # locations caught and not caught
M  <- pmax( 20, n*n2M )                                              # potential pop size
J  <- jmat[ plot, "2024" ]                                           # sampling bouts

mapCapture( n, A, X, Y )                                              # generate maps
```

```{r}
pvec <- logit.inv( mu.beta[1] + mu.beta[2]*d.vec^2 )
plot( d.vec, pvec, type = 'l', bty = 'n', las = 1, lwd = 2, xlab = 'Distance (m)', ylab = 'Probability p' )
```

```{r}
tmp <- fitModel( Y, J, M, X, A, n.mcmc, mu.beta, beta.tune, s2.beta, n.sim, mpath = mpath ) # MCMC 
out.1     <- tmp$out.1         # 1st stage output
out.2.P   <- tmp$out.2.P       # 2nd stage output
N.out.2.P <- tmp$N.out.2.P     # N estimates
fit       <- tmp$fit           # summary

par( bty = 'n', mar = c(2,4,1,1 ), omi = c(.5, .5, .5, .1) ) # MCMC chains
layout( matrix( 1:6,3,2) )
plot( out.1$beta.star[1,], type = "l", ylab = bquote( beta[0] ) , main = 'stage 1', xaxt = 'n' )
plot( out.1$beta.star[2,], type = "l", ylab = bquote( beta[1] ), xaxt = 'n' )
plot( out.1$psi.star, type="l", ylab = bquote( psi ) )
plot( out.2.P$beta.save[1,], type="l", lty=1, ylab = '', main = 'stage 2', xaxt = 'n' )
plot( out.2.P$beta.save[2,], type="l", lty=1, ylab = '', xaxt = 'n' )
plot( out.2.P$psi.save, type="l", ylab = '' )
mtext( 'MCMC iteration', side = 1, outer = T, line = 1 )
```

```{r}
wn <- which( startsWith( colnames( fit ), 'N.' ) )
fit[, wn]/.64
```

```{r}
posteriorPlots( out.2.P, N.out.2.P, n, M )

```

```{r}
wn <- which( startsWith( colnames( fit ), 'L.' ) )                # values
fit[, wn]

kernelPlot( P.2.P.mat, out.2.P )
p0  <- logit.inv( fit$B1.estimate )/2                             # add to plot
lci <- c( fit$L.CI_025, fit$L.estimate, fit$L.CI_975 )
for( j in 1:3)lines( lci[ c(j, j ) ], c( 0, p0 ), lty = 2 )
```

```{r}
# initialize to hold output:
recap <- matrix( 0, 1, length(capTimes) , dimnames = list( NULL, capTimes ) )
nvec  <- matrix( c( 0, 0, 0, NA ), 1, dimnames = list( NULL, nnames ) )
bvec  <- matrix( NA, 1, 8, dimnames = list( NULL, bnames ) )
pvec  <- matrix( NA, 1, 4, dimnames = list( NULL, pnames ) )
qvec  <- matrix( NA, 1, 4, dimnames = list( NULL, qnames ) )
lvec  <- matrix( NA, 1, 4, dimnames = list( NULL, dnames ) )
powerDetect <- NA

fit <- numeric( 0 )

for( s in 1:nspec ){
  
  ws    <- which( y$commonName == specNames[s] )    # this species 
  yspec <- y[ws , ]
  plots <- sort( unique( yspec$plot ) )             # plots where observed
  nplot <- length( plots )
  
  print( specNames[s] )
  print( table( yspec$plot, yspec$year ) )          # distribution of observations
  
  for( k in 1:nplot ){
    
    yall <- unique( y$year[ y$plot == plots[k] ] )  # years observed on this plot
    yrk  <- sort( unique( yspec$year ) )
    if( length( yrk ) == 0 )next
    
    for( j in 1:length(yall) ){
      
      df <- numeric( 0 )                            # initialize data.frame for this species/plot/yr
      yc <- as.character( yall[j] )
      J  <- jmat[ plots[k], yc ]                    # sample occasions
      wj <- which( yspec$year == yall[j] & yspec$plot == plots[k] )
      
      recap <- recap*0
      
      if( !yall[j] %in% yrk | length( wj ) == 0 ){          # placeholder if no captures
        
        df    <- data.frame( species = specNames[s], plot = plots[k], year = yall[j], J )
        df$n <- 0
        df   <- cbind( df, recap )
        df   <- cbind( df, nvec, bvec, pvec, qvec, powerDetect, lvec )
        
      }else{
        
        yj <- yspec[drop = F, wj, ]
        yt <- table( yj$trap, yj$id )                           # where observed
        id <- colnames( yt )
        n  <- length( id )
        Y  <- matrix( 0, L, n, dimnames = list( lnames, id ) )  # observed and not observed
        Y[ rownames( yt ), colnames( yt ) ] <- yt
        M  <- pmax( 20, n*n2M )
        
        yy <- table( colSums(Y) )
        names( yy ) <- paste( 'c', names(yy), sep = '') 
        yy <- yy[ names( yy ) %in% capTimes ]
        recap[ , names(yy) ] <- yy                              # times recaptured
        
        df <- data.frame( species = specNames[s], plot = plots[k], year = yall[j], J, n )
        df <- cbind( df, recap )
        
        tmp <- fitModel( Y, J, M, X, A, n.mcmc, mu.beta, beta.tune, s2.beta, n.sim )
        df  <- tmp$fit
      }
      fit <- rbind( fit, df )
      print( tail( fit ) )
    }
  }
}
```

```{r}
wn      <- which( startsWith( colnames( fit ), 'N.' ) )
density <- fit[, wn]/.3

py <- paste( fit$plot, fit$year, sep = '_' )
plotYears <- sort( unique( py ) )

yMean <- matrix( 0, length( plotYears ), nspec, dimnames = list( plotYears, specNames ) )
ySE   <- yMean
yMean[ cbind( py, fit$species) ] <- density$N.estimate
ySE[ cbind( py, fit$species) ]   <- density$N.SE
xdata <- data.frame( columnSplit( plotYears, '_' ) )
colnames( xdata ) <- c('plot','year')

plots <- sort( unique( xdata$plot ) )
pnum  <- c(1:length(plots) )
names( pnum ) <- plots

years <- sort( unique( xdata$year ) )

par( mfrow = c( nspec, 1 ), bty = 'n', mar = c(2,3,1,4), omi = c(.7,.5,.1,.1) )
cols <- c('#8c510a', '#01665e' )
for( j in 1:nspec ){
  
  wy   <- which( xdata$year == years[1] )
  nj   <- length( wy )
  xj   <- pnum[ xdata$plot[wy] ] - .1
  yj   <- yMean[ wy, specNames[j] ]
  ys   <- ySE[ wy, specNames[j] ]
  ylim <- c(0, max( yMean[ wy, specNames[j] ] + ySE[ wy, specNames[j] ] ) )
  
  plot( xj, yj, xlim = c(1, length( plots ) ), ylim = ylim, pch = 3, lwd = 2,
        xaxt = 'n', xlab = '', ylab = '', las = 1, col = cols[1] )
  segments( xj, yj - ys, xj, yj + ys, lwd = 2, col = cols[1] )
  axis(1, at = pnum, labels = F )
  
  wy <- which( xdata$year == years[2] )
  xj <- pnum[ xdata$plot[wy] ] + .1
  yj <- yMean[ wy, specNames[j] ]
  ys <- ySE[ wy, specNames[j] ]
  points( xj, yj, pch = 3, lwd = 2, col = cols[2] )
  segments( xj, yj - ys, xj, yj + ys, lwd = 2, col = cols[2] )
  title( specNames[j] )
}
axis( 1, at = pnum, labels = names( pnum ), las = 2 )
mtext('per ha', 2, outer = T )
legend( 'topleft', years, text.col = cols, bty = 'n' )
```
